 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-11-28 paper.Rmd
Typos fixed.

---
documentclass: jdssv
author:
  - name: Stuart Lee
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    email: \email{stuart.lee1@monash.edu}
  - name: Ursula Laa 
    affiliation: Monash University
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    # To add another line, use \AND at the end of the previous one as above
  - name: Dianne Cook
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    affiliation: Monash University
title:
  formatted: "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # If you use tex in the formatted title, also supply version without
  plain:     "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # For running headers, if needed
  short:     "\\pkg{liminal}: High-Dimensional Interactive Data Visualisation"
abstract: >
  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique. We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour. A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space. The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view. By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading. We display several case studies from both simulateions and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks. The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.
keywords:
  # at least one keyword must be supplied
  formatted: [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, "\\proglang{R}"]
  plain:     [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, R]
output: 
  bookdown::pdf_document2:
    template: template/jdssv_template.tex
    keep_tex: true
    toc: false
bibliography: liminal.bib
biblio-style: apalike
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.show = 'hold',
  dpi = 300,
  out.width = "\\textwidth",
  fig.path = here::here('img/')
)
```

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR method that aims to construct
an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
approximate the pair-wise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce].

@@

From the video linked in Figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (Figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (Figure \@ref(fig:gaussian)b).

@@

(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$. *(a)* The true data lies on 2-$d$ tree consisting of ten branches. This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce]. *(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view. *(c)* The default t-SNE view breaks the global structure of the tree. *(d)* Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.

@@

From the linked video, we see that the t-SNE view has been unable to 
recapitulate the topology of the tree - the backbone (blue) branch has been 
split into three fragments (Figure \@ref(fig:trees-01)a). We can see 
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video 
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (Figure \@ref(fig:trees-01)b).

@@

From the video linked in Figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(Figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(Figure \@ref(fig:mouse-01)d).

@@

From the video linked in Figure \@ref(fig:mouse-02), we see that the 
expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (Figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (Figure \@ref(fig:mouse-02)b).

>>>>

---
documentclass: jdssv
author:
  - name: Stuart Lee
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    email: \email{stuart.lee1@monash.edu}
  - name: Ursula Laa 
    affiliation: Monash University
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    # To add another line, use \AND at the end of the previous one as above
  - name: Dianne Cook
    address: |
      | Department of Econometrics and Business Statistics,
      | Monash University
    affiliation: Monash University
title:
  formatted: "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # If you use tex in the formatted title, also supply version without
  plain:     "Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings"
  # For running headers, if needed
  short:     "\\pkg{liminal}: High-Dimensional Interactive Data Visualisation"
abstract: >
  Non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) are ubiquitous in the natural sciences, however, the appropriate use of these methods is difficult because of their complex parameterisations; analysts must make trade-offs in order to identify structure in the visualisation of an NLDR technique. We present visual diagnostics for the pragmatic usage of NLDR methods by combining them with a technique called the tour. A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space. The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view. By linking the tour to an NLDR view, we can preserve global structure and through user interactions like linked brushing observe where the NLDR view may be misleading. We display several case studies from both simulations and single cell transcriptomics, that shows our approach is useful for cluster orientation tasks. The implementation of our framework is available as an \proglang{R} package called \pkg{liminal} available at \url{https://github.com/sa-lee/liminal}.
keywords:
  # at least one keyword must be supplied
  formatted: [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, "\\proglang{R}"]
  plain:     [dimensionality reduction, high-dimensional data, interactive graphics, t-SNE, grand tour, R]
output: 
  bookdown::pdf_document2:
    template: template/jdssv_template.tex
    keep_tex: true
    toc: false
bibliography: liminal.bib
biblio-style: apalike
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.show = 'hold',
  dpi = 300,
  out.width = "\\textwidth",
  fig.path = here::here('img/')
)
```

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct
an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
approximate the pair-wise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce].

@@

From the video linked in Figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (Figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (Figure \@ref(fig:gaussian)b).

@@

(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$. *(a)* The true data lies on a 2-$d$ tree consisting of ten branches. This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce]. *(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view. *(c)* The default t-SNE view breaks the global structure of the tree. *(d)* Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.

@@

From the linked video, we see that the t-SNE view has been unable to 
recapitulate the topology of the tree - the backbone (blue) branch has been 
split into three fragments (Figure \@ref(fig:trees-01)a). We can see 
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video 
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (Figure \@ref(fig:trees-01)b).

@@

From the video linked in Figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2, 4, 6, and 11 are
clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
the surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(Figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(Figure \@ref(fig:mouse-01)d).

@@

From the video linked in Figure \@ref(fig:mouse-02), we see that the 
expression of the marker genes, appear to separate
the previously clumped clusters 2, 4, 6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (Figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (Figure \@ref(fig:mouse-02)b).
